Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-19 paper/paper-v2.tex
paper rewrite section on types

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} For our purposes, we would rather like to identify a type by its cognitive make-up. A
type is a pair consisting of a lexicon and a pragmatic strategy. This is because we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use.

Lexica codify the truth-conditions of a language's expressions. A convenient way to represent
lexica is by $(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of
states of affairs (meanings) and $M$ a set of messages (forms available in the language). For
example, suppose that there are two relevant world states $\States = \set{\msome, \mall}$:
$\msome$ where Chris owns some but not all of Johnny Cash's albums and $\mall$ where Chris owns
them all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic names?}
\begin{align*}
  L_{4} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  L_{5} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level $n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?}
\begin{flalign}
&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}
\end{flalign}
According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
from above would then assign $\ssome$ a probability of one after hearing $\msome$, while a
literal interpreter with lexicon $L_5$ would assign $\ssome$ probability $.5$. The simplest
definition of a literal speaker would just be $S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha)$, but
the more complex formulation in (\ref{s:level0})

\begin{itemize}
\item no prob zero
\item $\alpha$ adjusts the level of avoidance of untrue signaling
\end{itemize}

 The behavior of literal speakers, given in
(\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which
speakers prefer one signal over another based on its expected communicative success.

\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}

Pragmatic behavior is similar to its literal counterparts. Their difference lies in that level
$n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely
relaying on their lexicon. That is, they reason about the way a rational level $n$ interlocutor
would use or interpret a message, and behave according to these expectations.  Speaker behavior
is further influenced by a soft-max parameter $\lambda$, $\lambda \geq 1$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices made in production are
more rational in that higher values lead to behavior that is increasingly in line with expected
utility maximization.

We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a
{\em type}. These are the basic units on which the model's dynamics operate.

>>>>

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} For our purposes here, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
  names?}
\begin{align*}
  L_{4} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  L_{5} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?}\mf{Literal speaker definition had $\alpha$ but that has no effect if we have Boolean
semantics; I left this out.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(L_{sm}) \label{s:level0}
\end{flalign}
According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; L_4) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $L_5$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; L_5) = 0.5$. A literal speaker chooses any true message
for a state with equal probability and prefers a true message over a false one. The formulation
in (\ref{s:level0}) allows also false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
necessary to guarantee a mutation matrix with only positive entries (see below). A literal
speaker with lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
\sall;L_4) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
  message for literal speakers is actually rather high. That's iffy because our pragmatic types
are able (given favorable parameter values) to perform much better. That means that we
implicitly have engineered in a possible fitness advantage of pragmatic types! Should we do
something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
only look at parameters for pragmatic types that make them not much better than the literal guys?}

Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}
As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ is here defined as $H_{n}(s|m;L)^\alpha$. The probability
$H_{n}(s|m;L)$ is the probability that the hearer will assign to or choose the correct
meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
speaker production behavior. To see this, notice that for two states and two messages (the
minimal scenario we will investigate), the possible values for $H_{0}(s|m;L)$ are restricted to
zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
probabilities to states after hearing messages:
\begin{align*}
  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & 0.5 & 0.5 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}
A level-1 speaker's utilities are then:
\begin{align*}
  \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
    \end{block}
  \end{blockarray}
\end{align*}
As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
informative message instead of a more informative one), from semantically deviant choices
(selecting a false message instead of a true one). If $\alpha > 1$ pragmatically deviant
choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
way around. In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically
or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
extension to environmental transmission noise in Section~XYZ. \mf{I would suggest to just give
  the intuitive summary here and have a little appendix paragraph with the example. What do you
say?}
